Synthesising Recursive Functions for First-Order Model Counting

### 2023-02-13 doc/paper/paper.tex
Improvements to the paper and more slides

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
% \newtheorem{example}{Example}
% \newtheorem{theorem}{Theorem}
\newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture}
\newtheorem{claim}{Claim}[theorem]
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

@@

@@

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator{\Preds}{Preds}
\SetKwProg{Fn}{Function}{:}{}
\SetKwFunction{identifyRecursion}{r}
\SetKwFunction{generateMaps}{genMaps}
\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother
\Crefname{algocf}{Algorithm}{Algorithms}
\crefname{line}{line}{lines}

@@

\section{Introduction}

% TODO: maybe mention the term 'data complexity'
% TODO: cite some works about synthesis

@@

TODO: cite \textsc{Magician}~\cite{DBLP:conf/aaai/VenugopalSG15} and
\textsc{Tuffy}~\cite{DBLP:journals/pvldb/NiuRDS11} (read the abstracts first,
look at the code)

@@

Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the instance size. The equivalent
notion in (W)FOMC is liftability. A (W)FOMC instance is \emph{(domain-)liftable}
if it can be solved in time polynomial in the size(s) of the
domain(s)~\cite{jaeger2012liftability}. Over more than a decade, many classes of
instances were shown to be
liftable~\cite{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} showed that the class
of all sentences of FOL with up to two variables (denoted \FOtwo{}) is liftable.
Then Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15} proved that there exists
a sentence with three variables for which FOMC is $\#\P_{1}$-complete (i.e.,
\FOthree{} is not liftable). Since these two results came out, most of the
research on (W)FOMC focused on developing faster solutions for the \FOtwo{}
fragment~\cite{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining
new liftable fragments. These fragments include \SFO{} and
\SRU{}~\cite{DBLP:conf/nips/KazemiKBP16},
\Uone{}~\cite{DBLP:conf/lics/KuusistoL18}, \FOtwo{} with tree
axioms~\cite{DBLP:conf/kr/BremenK21}, and \Ctwo{} (i.e., the two-variable
fragment with counting
quantifiers)~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. On
the empirical front, there are several open-source implementations of exact
WFOMC algorithms: \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11},
probabilistic theorem proving~\cite{DBLP:journals/cacm/GogateD16},
\textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16}, and
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}. Approximate counting is
supported by \textsc{ApproxWFOMC}~\cite{DBLP:conf/ijcai/BremenK20} as well as
\textsc{ForcLift}~\cite{DBLP:conf/uai/BroeckCD12} and probabilistic theorem
proving~\cite{DBLP:journals/cacm/GogateD16}.

@@

Once domains are mapped to finite sets and constants to elements in those sets,
a formula can be viewed as a set of conditions that the predicates (interpreted
as relations) have to satisfy. Hence, FOMC is the problem of counting the number
of combinations of relations that satisfy these conditions. To formalise this
idea, let $\sigma\colon \mathcal{D} \to \mathbb{N}_{0}$ be the \emph{domain size
  function} that maps each domain to a non-negative integer.

@@

\begin{example}
  Let $\phi$ be as in \cref{example:first} and let
  $\sigma(\Gamma) = \sigma(\Delta) = 2$. Then the right-hand side of
  \cref{eq:subset} becomes
  \[
    [\sigma(\Gamma)] \times [\sigma(\Delta)] = [2] \times [2] = \{\,(1, 1), (1, 2), (2, 1), (2, 2)\,\}
  \]
  There are $2^{4} = 16$ possible relations between $\Gamma$ and $\Delta$. Let
  us count how many of them satisfy the conditions imposed on predicate $p$. The
  empty relation does. All four relations of cardinality one do too. Finally,
  there are two relations of cardinality two that satisfy the conditions as
  well. Thus, the FOMC of $(\phi, \sigma)$ is 7. Incidentally, it counts partial
  injections. We will continue to use the problem of counting partial injections
  (and the formula from \cref{example:first} specifically) as the main running
  example throughout the paper.
\end{example}
% TODO: It might be useful to give one or two readable examples with interesting
% model counts. For example Px v Qx over domain n has 3^n solutions. I think
% The Frends smokers example has a more complicated solution you can find in one
% of the papers. It might motivate the framework generally speaking better
% although of course you can continue to use the same running example for the
% rest of your paper. Such examples are especially useful for readers who are
% not familiar but still need to get excited by the problem

@@

\begin{theorem}
  Any collection of functions returned by \textsc{Crane} can be evaluated in
  time polynomial in domain sizes.
\end{theorem}
\begin{proof}
  % 1) bound the number of possible arguments that any function can have
  % 2) prove that the computation time of any dynamic programming cell is always polytime (what about functions depending on other functions?))
  \begin{claim}
    Infinite recursion is impossible because that would necessitate a sequence
    of compilation rule applications that start with $\phi$ and end up with
    exactly $\phi$ as one of the outcomes. This is impossible because... [argue
    about formula length and the necessity to preserve domains and constants].
  \end{claim}
  \begin{claim}
    Argument values cannot increase from one function call to another.
  \end{claim}
  \begin{claim}
    Excluding functions, time to evaluate a function definition is polynomial
    (e.g., binomials, linear sums, powers, multiplication, addition). The number
    of operations within a definition is a constant that does not depend on
    domain sizes.
  \end{claim}
  \begin{claim}
    Describe a dynamic programming algorithm that:
    \begin{itemize}
      \item Gradually increases the sum of all domain sizes
      \item Computes and stores the values of all function calls whose arguments
            have that sum. Argue that there will always be an order in which
            that's possible.
    \end{itemize}
  \end{claim}
  NOTE: For proving that each cell is evaluated in polynomial time, I can just
  cite the ForcLift completeness (and domain recursion) paper.
\end{proof}

@@

We compare \textsc{Crane} and
\textsc{ForcLift}\footnote{\url{https://dtaid.cs.kuleuven.be/wfomc}}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
on their ability to count various kinds of functions. This class of instances is
chosen because of its simplicity and ubiquity and the inability of
state-of-the-art WFOMC algorithms to solve many such instances. Note that other
WFOMC algorithms---\textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16}, probabilistic
theorem proving~\cite{DBLP:journals/cacm/GogateD16}, and
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}---are unable to solve any of
the instances that \textsc{ForcLift} fails on. We begin by describing how such
function-counting problems can be expressed in FOL\@. \textsc{ForcLift} then
translates these sentences in FOL to formulas as defined in \cref{def:formula}.

>>>>

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
% \newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}
\newtheorem{fact}{Fact}
\newtheorem{proposition}{Proposition}
\newtheorem{conjecture}{Conjecture}
\newtheorem{claim}{Claim}[theorem]
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

@@

% TODO: use \textt for predicates (e.g., p, q), introduce the notation, use DeclareMathOperator

@@

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}

\DeclareMathOperator{\friends}{\texttt{friends}}
\DeclareMathOperator{\smokes}{\texttt{smokes}}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator{\Preds}{Preds}
\SetKwProg{Fn}{Function}{:}{}
\SetKwFunction{identifyRecursion}{r}
\SetKwFunction{generateMaps}{genMaps}

\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother
\Crefname{algocf}{Algorithm}{Algorithms}
\crefname{line}{line}{lines}

@@

\section{Introduction}\label{sec:introduction}

% TODO: cite some works about synthesis

@@

@@

Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the instance size. The equivalent
notion in (W)FOMC is liftability. A (W)FOMC instance is \emph{(domain-)liftable}
if it can be solved in time polynomial in the size(s) of the
domain(s)~\cite{jaeger2012liftability}. Over more than a decade, many classes of
instances were shown to be
liftable~\cite{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} showed that the class
of all sentences of FOL with up to two variables (denoted \FOtwo{}) is liftable.
Then Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15} proved that there exists
a sentence with three variables for which FOMC is $\#\P_{1}$-complete (i.e.,
\FOthree{} is not liftable). Since these two seminal results, most of the
research on (W)FOMC focused on developing faster solutions for the \FOtwo{}
fragment~\cite{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining
new liftable fragments. These fragments include \SFO{} and
\SRU{}~\cite{DBLP:conf/nips/KazemiKBP16},
\Uone{}~\cite{DBLP:conf/lics/KuusistoL18}, \FOtwo{} with tree
axioms~\cite{DBLP:conf/kr/BremenK21}, and \Ctwo{} (i.e., the two-variable
fragment with counting
quantifiers)~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. On
the empirical front, there are several implementations of exact WFOMC
algorithms: \textsc{Alchemy}~\cite{DBLP:journals/cacm/GogateD16},
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21},
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}, and
\textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16}. Approximate counting is
supported by \textsc{Alchemy}~\cite{DBLP:journals/cacm/GogateD16},
\textsc{ApproxWFOMC}~\cite{DBLP:conf/ijcai/BremenK20},
\textsc{ForcLift}~\cite{DBLP:conf/uai/BroeckCD12},
\textsc{Magician}~\cite{DBLP:conf/aaai/VenugopalSG15}, and
\textsc{Tuffy}~\cite{DBLP:journals/pvldb/NiuRDS11}.

% Magician: an approach to improve the efficiency of sampling and local search
% based MLN inference algorithms by utilising \#CSP algorithms. Tuffy: grounding
% and stochastic local search.

@@

Once domains are mapped to finite sets and constants to elements in those sets,
a formula can be viewed as a set of conditions that the predicates (interpreted
as relations) have to satisfy. Hence, FOMC is the problem of counting the number
of combinations of relations that satisfy these conditions.

\begin{example}\label{example:simple}
  Let $\Delta$ be a domain of size $n \in \mathbb{N}_{0}$. The model count of
  $\forall X \in \Delta\text{. } p(X) \lor q(X)$ is then $3^{n}$. Intuitively,
  since both predicates are of arity one, they can be interpreted as subsets of
  $\Delta$. Thus, the formula says that each element of $\Delta$ has to be
  included in $p$ or in $q$ or in both, i.e., there are three possibilities.
\end{example}

\begin{example}
  Consider a variant of the well-known `friends and smokers' example
  $\forall X, Y \in \Delta\text{.
  } \smokes(X) \land \friends(X, Y) \Rightarrow \smokes(Y)$. Letting
  $n \coloneqq |\Delta|$ as in \cref{example:simple}, the model count can be
  expressed as
  $\sum_{k=0}^{n} \binom{n}{k}2^{n^{2} - k(n-k)}$~\cite{DBLP:conf/kr/BroeckMD14}.
\end{example}

To formalise the idea of a model, let
$\sigma\colon \mathcal{D} \to \mathbb{N}_{0}$ be the \emph{domain size function}
that maps each domain to a non-negative integer.

@@

\begin{example}
  Let $\phi$ be as in \cref{example:first} and let
  $\sigma(\Gamma) = \sigma(\Delta) = 2$. Then the right-hand side of
  \cref{eq:subset} becomes
  \[
    [\sigma(\Gamma)] \times [\sigma(\Delta)] = [2] \times [2] = \{\,(1, 1), (1, 2), (2, 1), (2, 2)\,\}
  \]
  There are $2^{4} = 16$ possible relations between $\Gamma$ and $\Delta$. Let
  us count how many of them satisfy the conditions imposed on predicate $p$. The
  empty relation does. All four relations of cardinality one do too. Finally,
  there are two relations of cardinality two that satisfy the conditions as
  well. Thus, the FOMC of $(\phi, \sigma)$ is 7. Incidentally, it counts partial
  injections. We will continue to use the problem of counting partial injections
  (and the formula from \cref{example:first} specifically) as the main running
  example throughout the paper.
\end{example}

@@

@@

We compare \textsc{Crane} and
\textsc{ForcLift}\footnote{\url{https://dtaid.cs.kuleuven.be/wfomc}}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
on their ability to count various kinds of functions. This class of instances is
chosen because of its simplicity and ubiquity and the inability of
state-of-the-art WFOMC algorithms to solve many such instances. Note that other
WFOMC algorithms mentioned in \cref{sec:introduction} are unable to solve any of
the instances that \textsc{ForcLift} fails on. We begin by describing how such
function-counting problems can be expressed in FOL\@. \textsc{ForcLift} then
translates these sentences in FOL to formulas as defined in \cref{def:formula}.
